---
title: "Factor Analysis of IPIP Big Five Personality Data"
format:
  html:
    footnotes-hover: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## The Five-Factor Model of Personality (Big Five)

The Five-Factor Model (FFM), commonly referred to as the "Big Five" is
the prevailing empirical taxonomy in contemporary personality
psychology. Psychometric research have demonstrated that human
personality traits can be comprehensively organized into five broad
dimensions. Consequently, each individual can be mathematically
represented as a single point within this 5-dimensional continuous
space, where their specific coordinates correspond to their standing on
each respective trait.

These five latent constructs are:

-   **Extraversion:** The tendency to seek stimulation, be outgoing, and
    engage energetically with the external world.
-   **Neuroticism (or Emotional Instability):** The tendency to
    experience negative emotions, such as anxiety, anger, or depression,
    and to be highly responsive to stress.
-   **Agreeableness:** The tendency to be compassionate, cooperative,
    and empathetic rather than suspicious or antagonistic towards
    others.
-   **Conscientiousness:** The tendency to display self-discipline, act
    dutifully, and strive for achievement against measures or outside
    expectations.
-   **Openness to Experience:** The tendency to appreciate art, emotion,
    adventure, unusual ideas, and variety of experience.

In this analysis, we utilize data from the International Personality
Item Pool (IPIP), specifically a 50-item questionnaire designed to
measure these five distinct domains. The dataset analyzed here includes
responses from around 700 000 participants. Because these psychological
constructs are *latent*—meaning they cannot be observed or measured
directly—we must infer their presence and magnitude from the observable,
self-reported responses to the questionnaire items. To achieve this, we
rely on the mathematical framework of Exploratory Factor Analysis.[^1]

[^1]: It is sometimes more simply referred as Factor Analysis, however
    we prefer to be precise and distinguish Exploratory Factor Analysis
    from Confirmatory Factor Analysis (which will not be used here).

## The Mathematical Framework of Exploratory Factor Analysis (EFA)

Exploratory Factor Analysis (EFA) is a multivariate statistical
technique utilized to uncover the underlying latent structure of a
relatively large set of variables. Unlike Principal Component Analysis
(PCA), which is merely a data reduction technique aimed at maximizing
explained variance, EFA is a formal causal modeling technique. It
assumes that the observed correlations between variables are the result
of their shared reliance on a smaller number of unobserved continuous
latent variables (the factors). As a first approximation one can see EFA
as PCA's close cousin. In fact it has been shown empirically that
results from the PCA and the EFA are often very similar when the
communality is sufficiently high.

Let $X_1, X_2, \dots, X_p$ be our $p$ observed variables, which in this
context correspond to the $p=50$ individual questions of the survey. We
postulate that these variables are standardized (mean zero and variance
one). The common factor model asserts that each observed response $X_i$
can be expressed as a linear combination of $m$ underlying latent
factors $F_1, F_2, \dots, F_m$ (where $m < p$), plus a unique,
unobserved error term $\epsilon_i$. Variable by variable, the model is
formulated as:

$$X_i = l_{i1}F_1 + l_{i2}F_2 + \dots + l_{im}F_m + \epsilon_i \quad \text{for} \quad i = 1, \dots, p$$

Here, $l_{ij}$ represents the *factor loading*, which quantifies the
direct causal effect (or weight) of the $j$-th latent factor on the
$i$-th question.

Equivalently, letting $\mathbf{X} = [X_1, X_2, \dots, X_p]^T$ be the
random vector of the standardized observed responses, the model can be
expressed in its more compact matrix form:

$$\mathbf{X} = \mathbf{L}\mathbf{F} + \boldsymbol{\epsilon}$$

Where:

-   $\mathbf{L}$ is the $p \times m$ matrix of **factor loadings**
    ($l_{ij}$).
-   $\mathbf{F} = [F_1, F_2, \dots, F_m]^T$ contains the **common
    factors**, which are assumed to have a mean of zero, a variance of
    one and a correlation matrix $\boldsymbol{\Phi}$.[^2]
-   $\boldsymbol{\epsilon} = [\epsilon_1, \epsilon_2, \dots, \epsilon_p]^T$
    contains the **unique factors** (comprising both specific variance
    and measurement error). These are assumed to have a mean of zero and
    a covariance matrix $\boldsymbol{U^2}$.

[^2]: It is sometimes assumed that the common factors are uncorrelated,
    so that $\Phi = I$. In this case we get the simpler equation
    $\boldsymbol{R} =   \mathbf{L}\mathbf{L}^T+\boldsymbol{U^2}$.

A critical assumption of the EFA model is that the unique factors are
uncorrelated with each other and with the common factors. Therefore, the
*uniqueness matrix* $\boldsymbol{U^2}$ is a diagonal matrix. Under these
assumptions, the correlation matrix of the observed variables, denoted
as $\boldsymbol{R}$, can be decomposed into two distinct components:

$$\boldsymbol{R} = \mathbf{L}\boldsymbol{\Phi}\mathbf{L}^T + \boldsymbol{U^2}$$

Here, $\mathbf{L}\boldsymbol{\Phi}\mathbf{L}^T$ represents the variance
shared among the questions (the *communality*), and $\boldsymbol{U^2}$
represents the variance unique to each specific question. The
fundamental objective of EFA is to estimate the matrices $\mathbf{L}$
and $\boldsymbol{U^2}$ given the sample correlation or covariance matrix
of $\mathbf{X}$.

## Imports

```{r}
library(psych)       # factor analysis
library(ggplot2)     # data visualisation
library(pheatmap)    # heatmaps
library(tidyr)       # data reshaping
library(dplyr)       # data manipulation
library(gridExtra)   # combining multiple plots


options(digits = 5)
```

# Data Cleaning

The data file (around 400Mo) can be found on
[Kaggle](https://www.kaggle.com/datasets/tunguz/big-five-personality-test).

```{r}
data_raw <- read.csv("./IPIP-FFM-data-8Nov2018/data-final.csv", sep = "\t", na.strings = c("", " ", "NaN", "NA","NULL"))
```

```{r}
# Select columns to only keep the 50 items that measure the 5 personality traits.
# See codebook.txt for more details on all the columns.
data <- data_raw[, 1:50]

# Remove rows where IPC != 1, as these could be multiple submissions of the same person.
# The total number of rows in the original dataset is 1 015 341, and 696 845 after cleaning.
cat("Number of lines removed to avoid multiple submission:", sum(data_raw$IPC != 1), "\n")
data <- data[data_raw$IPC == 1, ]

# Remove rows with missing data
n_before <- nrow(data)
data <- data[complete.cases(data), ]
n_after  <- nrow(data)
cat("Number of lines removed to avoid missing data:", n_before - n_after, "\n")
cat("Number of lines :", n_after)
data_mat <- as.matrix(data)
head(data)
```

```{r}
# Computing polychoric correlations on large datasets is computationally heavy.
# We check if the matrix has already been saved to avoid recalculating it every time.

pmat_path <- "polychoric_matrix_IPIP-50.rds"

if (file.exists(pmat_path)) {
  cat("Loading saved polychoric matrix...\n")
  poly_mat <- readRDS(pmat_path)
} else {
  cat("Computing polychoric matrix (this may take a while)...\n")
  poly_mat_obj <- polychoric(data_mat, ML=FALSE) #ML=FALSE (default value) to use Olsson's two step procedure 
  poly_mat <- poly_mat_obj$rho
  
  # Save the matrix to a file for future runs
  saveRDS(poly_mat, pmat_path)
  cat("Polychoric matrix saved to", pmat_path, "\n")
}

trait_names <- c("Extraversion", "Neuroticism", "Agreeableness",
                 "Conscientiousness", "Openness")
```

# Factor Analysis

## Rotation and Interfactorial Correlations

We apply factor analysis to the data, to extract the Big Five
personality traits, with varimax rotation to make the traits more
interpretable.

We could have used non-orthogonal rotation methods, like oblimin, which
allow for correlated factors. It is commonly recommended to use oblique
rotation in psychometric analysis, as said in [Fabrigar et
al.(1999)](https://doi.org/10.1037/1082-989X.4.3.272):

> *For many constructs examined in psychology (e.g., mental abilities,
> personality traits, attitudes), there is substantial theoretical and
> empirical basis for expecting these constructs (or dimensions of these
> constructs) to be correlated with one another. Therefore, oblique
> rotations provide a more accurate and realistic representation of how
> constructs are likely to be related to one another.*

However, our goal here is not to explain the correlations between
factors, but to perform clustering on the factors, and we fear that
oblique rotation will deform the space in a way that will influence the
clustering. Additionally, [Costello and
Osborne(2005)](https://doi.org/10.7275/jyj1-4868) suggest that
orthogonal and oblique rotation often produce similar results, so we
will not lose much interpretability by using varimax rotation. The
results given in alternative B of this notebook agree with this
statement.

## Factor Analysis Extraction Method

`psych::fa()` with `fm="minres"` uses the minres method[^3] to perform
factor analysis, which does not assume that factors are normally
distributed, contrary to the maximum likelihood (ML) method.

[^3]: The minres method aim to minimise the square of the off diagonal
    coefficient of difference between the sample covariance matrix
    $\mathbf{S}$ and the model-implied covariance matrix i.e: \$
    \min\*{\*\\hat{\\mathbf{L}}} \sum{i \neq j} \left(
    \mathbf{S}*{ij}-(*\hat{\mathbf{L}}\hat{\mathbf{L}}\^T){ij}\right)\^2\$

Since the assumption that factors follow a Gaussian distribution
directly contradicts the assumption of the Gaussian mixture model we
will use later, we think it is preferable to use the ULS method instead
of the ML method. However, this does not seem to bother [Gerlach, M.,
Farb, B., Revelle, W. et
al.](https://doi.org/10.1038/s41562-018-0419-z), who use the ML method
before using GMM. For the sake of comparison, we also perform factor
analysis with the ML method in alternative A of this notebook, and the
results are very similar to those given by the ULS method.

## How Many Factors to Extract?

A common method to determine the number of factors to extract is to look
at the scree plot, which is a plot of the eigenvalues of the correlation
matrix of the data. The eigenvalues represent the amount of variance
explained by each factor. The scree plot shows a clear elbow at 5
factors, which suggests that we should extract 5 factors. This is
consistent with the fact that we are trying to extract the Big Five
personality traits.

```{r}
#| warning: false
# The ULS method has several names: minres, iterated principal factor method,
# or unweighted least squares method.
fa_result <- fa(poly_mat, nfactors = 5, rotate = "varimax", fm = "minres", n.obs = nrow(data_mat), scores="Anderson")

data_fa <- as.data.frame(factor.scores(data_mat, fa_result)$scores)

# Eigenvalues of the original (unreduced) correlation matrix
eigenvalues <- fa_result$values
cat("Eigenvalues:", eigenvalues, "\n")

# Plot the ten first eigenvalues in a scree plot
top_10_eigenvalues <- eigenvalues[1:10]
scree_df <- data.frame(factor = 1:10, eigenvalue = top_10_eigenvalues)

ggplot(scree_df, aes(x = factor, y = eigenvalue)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "blue", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Scree Plot", x = "Factor Number", y = "Eigenvalue") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

Looking at the scree plot, it is not clear that we should extract five
factors; the solution with six factors also seems plausible. We will
look at Velicer's Minimal Average Partial (MAP) criterion, which is
commonly considered to be one of the best criterion for choosing how
much factor we should extract.

### Velicer's MAP criterion

Before defining the MAP criterion, we first need to define the partial
correlation matrix.

Let $X$ be a centered random vector of dimension $(p + m) \times 1$. We
partition $X$ into two subvectors: $X_1$ of dimension $p \times 1$ and
$X_2$ of dimension $m \times 1$, such that:

$$X = \begin{bmatrix} X_1 \\ X_2 \end{bmatrix}$$

Let $C:= \text{Cov}(X)$ . Partitioning $C$ conformably with $X$ yields:

$$C = \begin{bmatrix} C_{11} & C_{12} \\ C_{12}^T & C_{22} \end{bmatrix}$$
where $C_{11} = \text{Cov}(X_1)$, $C_{22} = \text{Cov}(X_2)$, and
$C_{12} = \text{Cov}(X_1, X_2)$. We assume that $C_{22}$ is nonsingular.

To partial out the effect of $X_2$ from $X_1$, we consider the best
linear predictor of $X_1$ based on $X_2$. This is derived from the
population linear regression of $X_1$ on $X_2$. The matrix of regression
coefficients is given by $C_{12}C_{22}^{-1}$[^4].

[^4]: For those who are not familiar with this last expression, we can
    give an equivalent expression in the usual (more restrictive)
    multiple linear regression framework:

    Let's assume that we have a uniform probability on the sample space
    $\Omega = \{1,\dots,n\}$ and that for all $i \in \Omega$,
    $X_2(i) = \mathbb{X}_{i, \cdot}^T$ and $X_1(i)=\mathbb{Y}_i$ :
    $C_{22}$ would be $\frac{1}{n}\mathbb{X}^T\mathbb{X}$, and $C_{21}$
    would be $\frac{1}{n}(\mathbb{X}^T \mathbb{Y})$, so that we get
    $\hat{\beta} = (\frac{1}{n} \mathbb{X}^T \mathbb{X})^{-1} (\frac{1}{n} \mathbb{X}^T \mathbb{Y}) = {C}_{22}^{-1} {C}_{21}$,
    and so $\hat{X}_1 = (C_{12}C_{22}^{-1}) X_2$ translates into the
    well known formula $\hat{X}_1 = \hat{\beta}^T X_2$.

We now define the residual vector $e$, which represents the portion of
$X_1$ that is orthogonal to (and thus uncorrelated with) the vector
space spanned by $X_2$:

$$e = X_1 - C_{12}C_{22}^{-1}X_2$$

The matrix of partial variances and covariances of $X_1$ given $X_2$,
denoted $C_{11}^*$, is defined as the covariance matrix of the residual
vector $e$. By applying the bilinear properties of the covariance
operator, we compute $\text{Cov}(e)$:

$$C_{11}^* = \text{Cov}(X_1 - C_{12}C_{22}^{-1}X_2)$$
$$C_{11}^* = \text{Cov}(X_1) + \text{Cov}(C_{12}C_{22}^{-1}X_2) - \text{Cov}(X_1, C_{12}C_{22}^{-1}X_2) - \text{Cov}(C_{12}C_{22}^{-1}X_2, X_1)$$
Applying the property $\text{Cov}(AY) = A\text{Cov}(Y)A'$ for a constant
matrix $A$ and random vector $Y$:

$$C_{11}^* = C_{11} + (C_{12}C_{22}^{-1})\text{Cov}(X_2)(C_{12}C_{22}^{-1})' - \text{Cov}(X_1, X_2)(C_{12}C_{22}^{-1})' - (C_{12}C_{22}^{-1})\text{Cov}(X_2, X_1)$$
Substitute the block matrices $C_{22} = \text{Cov}(X_2)$ and
$C_{12} = \text{Cov}(X_1, X_2)$. Note that
$\text{Cov}(X_2, X_1) = C_{12}'$ and $(C_{22}^{-1})' = C_{22}^{-1}$ due
to symmetry:

$$C_{11}^* = C_{11} + C_{12}C_{22}^{-1}C_{22}C_{22}^{-1}C_{12}' - C_{12}C_{22}^{-1}C_{12}' - C_{12}C_{22}^{-1}C_{12}'$$
Since $C_{22}C_{22}^{-1} = I$, the expression simplifies to:

$$C_{11}^* = C_{11} + C_{12}C_{22}^{-1}C_{12}' - 2C_{12}C_{22}^{-1}C_{12}'$$
$$C_{11}^* = C_{11} - C_{12}C_{22}^{-1}C_{12}'$$

A correlation matrix is derived by scaling a covariance matrix by the
inverse of the standard deviations of its variables. Let $D$ be a
diagonal matrix containing the main diagonal elements of the partial
covariance matrix $C_{11}^*$:

$$D = \text{Diag}(C_{11}^*)$$The diagonal elements of $D$ are the
partial variances. Consequently, $D^{-1/2}$ is a diagonal matrix
containing the reciprocals of the partial standard deviations.
Pre-multiplying and post-multiplying $C_{11}^*$ by this matrix
standardizes the partial covariances, yielding the partial correlation
matrix $R_{11}^*$:

$$R_{11}^* = D^{-1/2}C_{11}^*D^{-1/2}$$Substituting the derived
expression for $C_{11}^*$, we obtain
$R_{11}^* = D^{-1/2}(C_{11} - C_{12}C_{22}^{-1}C_{12}')D^{-1/2}$.

To connect the general partial correlation framework to the EFA model,
the notations translate as follows:

-   $X_1 \equiv \mathbf{X}$: The vector of observed variables.

-   $X_2 \equiv \mathbf{F}$: The vector of latent factors (or extracted
    principal components).

-   $C^*_{11} \equiv \boldsymbol{U^2}$: The residual covariance matrix
    corresponds to the diagonal uniqueness matrix.

-   $R^*_{11} \equiv \mathbf{I}$ (Theoretically): The partial
    correlation matrix of the observed variables after partialling out
    the factors. As the standardized version of the diagonal
    $\boldsymbol{U^2}$ matrix, a perfectly specified theoretical model
    yields the identity matrix.

The Minimum Average Partial (MAP) criterion is just the sum of squared
off diagonal elements of the *empirical* partial correlation matrix
$\hat{R}_{11}^*$, denoted here by $f_m$. It is supposed to be minimal
when the number of extracted factors $m$ attains its optimal values.

Here is another explanation of the rational behind this criterion from
[Velicer's original 1976 paper](https://doi.org/10.1007/BF02293557):

> In order to better understand why the value $f_m$ first decreases and
> then increases, consider the general form of a partial correlation
> equation $$
> \begin{equation*}
> r_{i j, y}=\frac{r_{i j}-r_{j y} r_{i y}}{\left(\left(1-r_{j y}{ }^2\right)\left(1-r_{i y}{ }^2\right)\right)^{1 / 2}}
> \end{equation*}
> $$ where $i, j$ represent any two of the $p$ observed variables and
> $y$ represents a \[factor\]. With no loss of generality, we assume
> $r_{i j}, r_{i j}$, and $r_{i y}$ to be positive. The value of the
> partial correlation will decrease as long as the numerator decreases
> faster than the denominator. The value of the partial correlation will
> increase whenever the reverse is true; for example, when $r_{iy}$ is
> large and $r_{j y}$ is small. *Such a situation would occur if a
> \[factor\] has a high correlation with only one variable and near zero
> correlations with the others (commonly referred to as a "specific")*.
> This explanation can be extended inductively to the case where more
> than one \[factor\] has been extracted by using $k$-th order partials
> calculated from ( $k-1$-th) order partials. Using this stopping rule,
> the first $m$ \[factors\] could be referred to as "common
> \[factors\]". We would expect that two or more variables would have
> "high" loadings on these \[factors\]. The remaining $m+1$ to $p$
> \[factors\] may include some that have adequate reliability. However,
> these will be "unique" \[factors\] (i.e., having "high" loadings on
> only one of the original variables).

In other terms, as you extract "common factors" (which relate to
multiple variables), the remaining correlations between those variables
successfully decrease. But if you extract a "unique factor" (which
relates strongly to only *one* variable), the mathematical formula
shrinks the denominator more than the numerator, which artificially
inflates the remaining correlations. Therefore, the overall value $f_m$
drops while you are finding useful common factors, and starts to rise as
soon as you begin extracting factors specific to single variables.

### Very Simple Structure criterion

Most people, when interpreting a factor solution, will highlight the
large (salient) loadings and ignore the small loadings. That is, they
are interpreting the factor matrix as if it had simple structure. How
well does this simplified matrix reproduce the original matrix ? To
quantify this, let $c$ be the complexity (number of non zero loadings)
of an item and $max_c$ be the greatest (absolute) $c$ loadings for an
item, then find the Very Simple Structure matrix, $\hat{\mathbf{S}}_c$,
where
$$ \hat{s}_{cij} = \begin{cases} \hat{l}_{ij} & \text{if } \hat{l}_{ij} \in max_c(\hat{l}_{i.}) \\ 0 & \text{otherwise} \end{cases} $$Then
let\
$$ \hat{\mathbf{R}}_{sc}^* = \hat{\mathbf{R}} - \hat{\mathbf{S}}_c\hat{\mathbf{S}}_c^T $$and\
$$ VSS_c = 1 - \frac{\mathbf{1}{\hat{\mathbf{R}}_{sc}^{*\odot 2}}\mathbf{1}' - tr(\hat{\mathbf{R}}_{sc}^*)}{\mathbf{1}\hat{\mathbf{R}}^{\odot 2}\mathbf{1}' - tr(\hat{\mathbf{R}})} $$The
VSS criterion for a complexity $c$ is 1 - the squared residual
correlations divided by the squared observed correlations, where the
residuals are found by the “simplified” factor equation
$\hat{\mathbf{R}}_{sc}^* = \hat{\mathbf{R}} - \hat{\mathbf{S}}_c\hat{\mathbf{S}}_c^T$.
[^5]

[^5]: This presentation of the VSS criterion comes from the
    [lectures](https://www.personality-project.org/r/book/Chapter6.pdf)
    of one its authors, W.Revelle (who is also the author of the psych R
    package we are using here).

The VSS criterion is less used than the MAP in the literature. In fact
the VSS criterion makes the strong assumption that the loadings of
variable on factors are simple, which is not always the case. Our
personal preference goes to the MAP criterion, which seems more
objective.

Here are now the results of both MAP and VSS criterion.

```{r}
#| warning: false
# Run MAP and VSS test on the polychoric correlation matrix
n_of_factors_diagnostic <- VSS(poly_mat, n.obs = nrow(data_mat), n = 10, plot = FALSE)
summary(n_of_factors_diagnostic)
```

The MAP criterion advises us to extract six factors. In fact six factor
model have largely been discussed in the literature, such as the HEXACO,
which is more or less the Big 5 with the addition of a honesty-humility
trait. However, since our data have been specifically designed to
estimate the Big 5 model, we will stick to a five factor model here.

```{r}
# Plot the loadings of the 50 items on the 5 personality traits as a heatmap.
# NOTE: The factor order below is inferred by inspecting which group of items
# loads most strongly on each factor (see the loading heatmap).

data_fa  <- data_fa[, c(2,1,3,5,4)]
fa_result$loadings <- fa_result$loadings[, c(2,1,3,5,4)]

# fa$loadings is items × factors; transpose to get traits × items
weights <- t(as.matrix(fa_result$loadings))


rownames(weights) <- trait_names

coolwarm <- colorRampPalette(c("#3B4CC0", "white", "#B40426"))(200)
lim      <- max(abs(weights))
breaks   <- seq(-lim, lim, length.out = 201)

pheatmap(
  weights,
  color        = coolwarm,
  breaks       = breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main         = "Weights of the 50 items on the 5 personality traits",
  fontsize      = 8,
  fontsize_row  = 9
)
```

```{r}
#| fig-width: 15
#| fig-height: 4
# From the previous plot, we can infer that factors are ordered as listed in trait_names.
# Assign the Big Five trait names as column names.
colnames(data_fa) <- trait_names


data_fa_long <- pivot_longer(data_fa, cols = everything(),
                             names_to = "trait", values_to = "score")
data_fa_long$trait <- factor(data_fa_long$trait)

ggplot(data_fa_long, aes(x = score)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, alpha = 0.6, color = NA) +
  geom_density(color = "darkred", linewidth = 0.8) +
  facet_wrap(~trait, nrow = 1, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(x = "Score", y = "Density")
```

```{r}
write.csv(data_fa, "IPIP-50_FFM_factor_scores.csv", row.names = FALSE)
```

# Alternative A: Comparison of ULS vs ML Extraction Methods for Factor Analysis

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 15
# Perform factor analysis with the maximum likelihood (ML) method,
# for comparison with the ULS method.
fa_ml     <- fa(poly_mat, nfactors = 5, rotate = "varimax", fm = "ml", n.obs = nrow(data_mat))
data_fa_ml <- as.data.frame(factor.scores(data_mat, fa_ml)$scores)

# Exchange column to properly match with traits
data_fa_ml  <- data_fa_ml[, c(1, 2, 3, 5, 4)]
fa_ml$loadings <- fa_ml$loadings[, c(1, 2, 3, 5, 4)]

colnames(data_fa_ml) <- trait_names
loadings_ml <- as.matrix(fa_ml$loadings)

weights_ml <- t(loadings_ml)

# Loading matrix comparison plots
breaks_full <- seq(-1, 1, length.out = 201)

p1 <- pheatmap(weights_ml, color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Factor loading (ML Method)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)
p2 <- pheatmap(weights,    color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Factor loading (ULS Method)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)
diff_ml <- unclass(weights)- unclass(weights_ml) #unclass to avoid the rounding of "loadings" class
p3 <- pheatmap(diff_ml,    color = coolwarm, breaks = breaks_full,
               cluster_rows = FALSE, cluster_cols = FALSE,
               main = "Difference (ML - ULS)",
               fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p1$gtable, p2$gtable, p3$gtable, ncol = 3,
             widths = c(1, 1, 1))

cat("Difference between the loadings of the ML method and the ULS method:\n")
diff_ml
```

```{r}
#| fig-height: 4
#| fig-width: 15
# Histogram comparison: ML vs ULS distributions for each trait

data_compare <- bind_rows(
  pivot_longer(data_fa_ml, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "ML"),
  pivot_longer(data_fa, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "ULS")
)
data_compare$trait <- factor(data_compare$trait)

ggplot(data_compare, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("ML" = "royalblue", "ULS" = "darkred")) +
  scale_color_manual(values = c("ML" = "royalblue", "ULS" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

We can see that the loadings and distributions of the factors with the
ULS method are very similar to those given by the factor analysis with
the ML method, which suggests that the choice of extraction method does
not have a strong influence on the results. However, we think it is
preferable to use the ULS method, since it does not assume that factors
are normally distributed, which is more consistent with the assumptions
of the Gaussian mixture model we will use later.

# Alternative B: Comparison of Varimax vs Oblimin Rotation Methods for Factor Analysis

In the main part of the notebook, we use orthogonal rotation (varimax)
to make the factors more interpretable. However, we could have used
oblique (oblimin) rotation, which allows for correlated factors. We will
see that the results given by the oblimin rotation are very similar to
those given by the varimax rotation, which suggests that the choice of
rotation method does not have a strong influence on the results.

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 15
# Perform factor analysis with the oblimin rotation method,
# for comparison with the varimax rotation used above.
# scores="tenBerge" to preserve correlation
fa_obli      <- fa(poly_mat, nfactors = 5, rotate = "oblimin", fm = "minres", n.obs = nrow(data_mat), scores="tenBerge")
data_fa_obli <- as.data.frame(factor.scores(data_mat, fa_obli)$scores)
```

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 15

# Exchange column to properly match with traits
data_fa_obli  <- data_fa_obli[, c(1, 2, 3, 5, 4)]
fa_obli$loadings <- fa_obli$loadings[, c(1, 2, 3, 5, 4)]

# From our (not shown) observations, factors are ordered as follows:
colnames(data_fa_obli) <- trait_names

loadings_obli  <- as.matrix(fa_obli$loadings)

weights_obli <- t(loadings_obli)
rownames(weights_obli) <- trait_names

# Loading matrix comparison plots
p_obli <- pheatmap(weights_obli, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Oblique Rotation)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)
p_vari <- pheatmap(weights,      color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Orthogonal Rotation)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)
diff_obli <- unclass(weights_obli) - unclass(weights)
p_diff <- pheatmap(diff_obli,   color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Difference (Oblique - Orthogonal)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p_obli$gtable, p_vari$gtable, p_diff$gtable, ncol = 3,
             widths = c(1, 1, 1))
```

## Correlation Between the Factors Obtained with the Oblimin Rotation

Unlike the factors obtained with the varimax rotation, the factors
obtained with the oblimin rotation are correlated, as expected. However,
the correlations between factors are not very strong, which suggests
that the factors are still relatively independent from each other. This
is consistent with the fact that the results given by the oblimin
rotation are very similar to those given by the varimax rotation. We
give here a comparison of the correlation matrix we found with the one
given by Van der Linden et al. in their [2010
meta-analysis](https://doi.org/10.1016/j.jrp.2010.03.003).

```{r}
#| fig-height: 4
#| fig-width: 11
# Correlation matrix of the 5 personality traits obtained with the oblique rotation
corr_obli <- fa_obli$Phi

rownames(corr_obli) <- colnames(corr_obli) <- trait_names

# (Uncorrected) correlation matrix from the meta-analysis of Van der Linden et al. (2010)
corr_meta <- matrix(c(
   1.00, -0.26,  0.18,  0.21,  0.31,
  -0.26,  1.00, -0.26, -0.32, -0.12,
   0.18, -0.26,  1.00,  0.31,  0.14,
   0.21, -0.32,  0.31,  1.00,  0.14,
   0.31, -0.12,  0.14,  0.14,  1.00
), nrow = 5, byrow = TRUE,
   dimnames = list(trait_names, trait_names))

sd_meta <- matrix(c(
  0.00, 0.11, 0.15, 0.15, 0.12,
  0.11, 0.00, 0.14, 0.18, 0.12,
  0.15, 0.14, 0.00, 0.14, 0.12,
  0.15, 0.18, 0.14, 0.00, 0.15,
  0.12, 0.12, 0.12, 0.15, 0.00
), nrow = 5, byrow = TRUE)

# Build annotation matrices
annot_obli <- matrix(sprintf("%.2f", corr_obli), nrow = 5,
                     dimnames = dimnames(corr_obli))
annot_meta <- matrix("", nrow = 5, ncol = 5, dimnames = dimnames(corr_meta))
for (i in 1:5) {
  for (j in 1:5) {
    if (i == j) {
      annot_meta[i, j] <- sprintf("%.2f", corr_meta[i, j])
    } else {
      annot_meta[i, j] <- sprintf("%.2f \n(+/-%.2f)", corr_meta[i, j], sd_meta[i, j])
    }
  }
}

# Show the two correlation matrices side by side for comparison
p_corr_obli <- pheatmap(
  corr_obli, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_obli, number_format = "%.2f",
  main = "Correlation Matrix (Oblique Rotation)",
  fontsize = 9, silent = TRUE
)
p_corr_meta <- pheatmap(
  corr_meta, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_meta,
  main = "Correlation Matrix (Meta-Analysis)",
  fontsize = 9, silent = TRUE
)

grid.arrange(p_corr_obli$gtable, p_corr_meta$gtable, ncol = 2,
             widths = c(1, 1))
```

```{r}
#| fig-height: 4
#| fig-width: 11
# Histogram comparison: oblimin vs varimax distributions for each trait
data_compare_obli <- bind_rows(
  pivot_longer(data_fa_obli, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "oblimin (oblique)"),
  pivot_longer(data_fa, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "varimax (orthogonal)")
)
data_compare_obli$trait <- factor(data_compare_obli$trait)

ggplot(data_compare_obli, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("oblimin (oblique)" = "royalblue", "varimax (orthogonal)" = "darkred")) +
  scale_color_manual(values = c("oblimin (oblique)" = "royalblue", "varimax (orthogonal)" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

# Alternative C: Pearson vs polychoric matrix for factor extraction

## Polychoric correlation matrix

### The trouble with Likert scales

Exploratory factor analysis is based on a correlation matrix: from a
correlation matrix, it infers the underlying factors that explain the
correlations between the items. The core assumption of our study is that
the responses to the questionnaire are strongly determined by some
underlying *continuous* factors. So it seems intuitive that each person
has an "internal" *continuous* response to each question that is a
linear combination of the underlying (continuous) factors plus a random
noise. This "internal" response is then discretized into an ordinal
Likert scale when answering the questionnaire. In a perfect world, we
would have direct access to these "internal" continuous responses, which
are more reflective of the personality of the participant, and then we
would perform the factor analysis on the correlations of these
"internal" continuous responses. But we only have access to the Likert
scales, which correlation underestimate the "true" underlying
correlation, as we can see in the following theorem:

> **Theorem:** If $(Z_1, Z_2)$ have a joint bivariate normal
> distribution with correlation coefficient $\rho$, then for any
> Borel-measurable functions $f$ and $g$ such that $f(Z_1)$ and $g(Z_2)$
> possess finite variances, the absolute correlation between the
> transformed variables is bounded by the absolute value of $\rho$:
> $$|\text{corr}(f(Z_1), g(Z_2))| \leq |\rho|$$ The equality only
> happens when $f=g=id$.

> **Proof:** The proof rely on the fact that we can express $L^2$ random
> variables (w.r.t. Gaussian measure) as series of standardized
> Hermite-Tchebycheff polynomial, which are orthogonal w.r.t. Gaussian
> measure. See [H.O. Lancaster(1957)](https://doi.org/10.2307/2333274)
> for the full proof.

Now if we assume that $Z_1$ and $Z_2$ are Gaussian, we can define $L_1$
and $L_2$ as the observed ordinal Likert variables. These discrete
variables are created by partitioning the continuous variables $Z_1$ and
$Z_2$ using a set of thresholds (e.g., $L_1 = i$ if
$\alpha_{i-1} < Z_1 \leq \alpha_i$, and $L_2 = j$ if
$\beta_{j-1} < Z_2 \leq \beta_j$). Since $L_1$ and $L_2$ are created
through step functions of the underlying continuous variables (i.e.,
$L_1 = f(Z_1)$ and $L_2 = g(Z_2)$), it rigorously follows that the
correlation between the discrete responses is strictly bounded by the
correlation of the "internal" continuous variables:

$$|\text{corr}(L_1, L_2)| < |\rho|$$

This is why Likert-scale questionnaires tend to underestimate the "true"
correlation between the questions. From this example, it seems very
plausible that the naive use of Pearson correlation on Likert scale can
give poor results. One way of addressing this issue is to use a
polychoric correlation matrix.

### Polychoric correlation estimation

Given what we have said above, if we use the Pearson correlation matrix,
we will obtain an estimation of the correlation of the Likert discrete
variables, not the correlation of the underlying continuous ones. So we
need a way to infer the correlation of the underlying continuous
variables given their discretized versions. This is exactly what
polychoric correlation is supposed to do.

To compute the polychoric correlation, we assume that the underlying
continuous latent variables $(Z_1, Z_2)$ follow a standard bivariate
normal distribution with correlation $\rho$. The theoretical probability
$\pi_{ij}$ of an observation falling into the $i$-th category for $L_1$
and the $j$-th category for $L_2$ is the following:

$$\pi_{ij} = \Phi_2^{}(\alpha_i, \beta_j; \rho) - \Phi_2(\alpha_{i-1}, \beta_j; \rho) - \Phi_2(\alpha_i, \beta_{j-1}; \rho) + \Phi_2(\alpha_{i-1}, \beta_{j-1}; \rho)$$

where $\Phi_2$ is the cumulative distribution function (CDF) of the
standard bivariate normal distribution.

To estimate these parameters from an observed dataset where $n_{ij}$
represents the frequency of responses in cell $(i, j)$, we use Maximum
Likelihood Estimation (MLE). The log-likelihood function to maximize is
defined as:

$$\ell(\rho, \mathbf{\alpha}, \mathbf{\beta}) = \sum_{i} \sum_{j} n_{ij} \ln[\pi_{ij}(\rho, \alpha, \beta)]$$

**Olsson's Two-Step Method** Maximizing this log-likelihood for all
parameters ($\rho$, $\alpha$, $\beta$) simultaneously is computationally
expensive. To solve this, the standard approach relies on [Olsson's
(1979)](https://doi.org/10.1007/BF02296207) two-step procedure, which
consists of first estimating the thresholds $\alpha$ and $\beta$ using
only the marginal distributions, and then estimating the correlation
coefficient $\rho$. Despite the theoretical non-optimality of this
procedure, the gain in computational efficiency outweighs the loss in
precision, as reported in Olsson's article. However, it is still
possible to use a full maximum likelihood estimation for maximum
statistical efficiency. In this code, we will use Olsson's two step
method (parameter `ML=FALSE` in the `polychoric` function from `psych`
package).

### Empirical evidence of the superiority of polychoric correlation matrix in EFA

The methodological importance of this choice is highlighted by
[Holgado–Tello et al (2008)](https://doi.org/10.1007/s11135-008-9190-y).
The authors demonstrate that when data are ordinal, polychoric
correlations provide a more accurate and stable reproduction of the
original factor structure compared to Pearson correlations, which can
lead to biased loadings and incorrect factor counts.

```{r}
#| warning: false
#| message: false
#| fig-height: 4
#| fig-width: 15

# Perform factor analysis with the oblimin rotation method using Pearson correlations.
# NOTE: The default was set to polychoric at the beginning of the notebook for all main analyses.
fa_pearson      <- fa(data_mat, nfactors = 5, rotate = "oblimin", fm = "minres",
                   cor = "cor", scores="tenBerge")
data_fa_pearson <- as.data.frame(fa_pearson$scores)
loadings_pearson  <- as.matrix(fa_pearson$loadings)

# Apply the trait names to the columns and rows
colnames(data_fa_pearson) <- trait_names

weights_pearson <- t(loadings_pearson)
rownames(weights_pearson) <- trait_names

# Loading matrix comparison plots
p_poly <- pheatmap(weights_obli, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Polychoric Oblimin)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

p_pearson <- pheatmap(weights_pearson, color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Factor loading (Pearson Oblimin)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

# Calculate difference using unclass() to prevent the "loadings" rounding issue
diff_pearson <- unclass(weights_pearson) - unclass(weights_obli)

p_diff <- pheatmap(diff_pearson,   color = coolwarm, breaks = breaks_full,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "Difference (Pearson - Polychoric)",
                   fontsize = 7, fontsize_row = 8, silent = TRUE)

grid.arrange(p_poly$gtable, p_pearson$gtable, p_diff$gtable, ncol = 3,
             widths = c(1, 1, 1))
```

```{r}
#| fig-height: 4
#| fig-width: 15
# Histogram comparison: Pearson vs Polychoric correlation matrix for factor analysis for each trait

data_compare_pearson <- bind_rows(
  pivot_longer(data_fa_pearson, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "Pearson"),
  pivot_longer(data_fa_obli, cols = everything(),
               names_to = "trait", values_to = "score") %>%
    mutate(method = "Polychoric")
)

# Convert trait to factor to control facet order if needed
data_compare_pearson$trait <- factor(data_compare_pearson$trait)

ggplot(data_compare_pearson, aes(x = score, color = method, fill = method)) +
  
  geom_density(alpha = 0.4, linewidth = 0.8) +
  
  facet_wrap(~trait, nrow = 1, scales = "free") +
  
  scale_fill_manual(values  = c("Pearson" = "royalblue", "Polychoric" = "darkred")) +
  scale_color_manual(values = c("Pearson" = "royalblue", "Polychoric" = "darkred")) +
  
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom") +
  labs(x = "Score", y = "Density")
```

```{r}
#| fig-height: 5
#| fig-width: 16
#| warning: false

corr_poly <- fa_obli$Phi

corr_pearson <- fa_pearson$Phi

annot_poly <- matrix(sprintf("%.2f", corr_poly), nrow = 5)
annot_pearson <- matrix(sprintf("%.2f", corr_pearson), nrow = 5)

rownames(corr_poly) <- colnames(corr_poly) <- trait_names
rownames(corr_pearson) <- colnames(corr_pearson) <- trait_names

p_corr_poly <- pheatmap(
  corr_poly, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_poly, number_format = "%.2f",
  main = "Factor Correlation (Polychoric Oblimin)",
  fontsize = 9, silent = TRUE
)

p_corr_pearson <- pheatmap(
  corr_pearson, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_pearson, number_format = "%.2f",
  main = "Factor Correlation (Pearson Oblimin)",
  fontsize = 9, silent = TRUE
)

p_corr_meta_comp <- pheatmap(
  corr_meta, color = coolwarm, breaks = breaks_full,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = annot_meta,
  main = "Factor Correlation (Meta-Analysis)",
  fontsize = 9, silent = TRUE
)

grid.arrange(p_corr_poly$gtable, p_corr_pearson$gtable, p_corr_meta_comp$gtable, 
             ncol = 3, widths = c(1, 1, 1))
```

```{r}
#| fig-height: 5
#| fig-width: 16
#| warning: false

pearson_mat <- cor(data_mat)


diff_item_mat <- poly_mat - pearson_mat

lim_diff <- max(abs(diff_item_mat))
breaks_diff <- seq(-lim_diff, lim_diff, length.out = 201)

p_item_pearson <- pheatmap(
  pearson_mat, 
  color = coolwarm, 
  breaks = breaks_full,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Item Correlation (Pearson)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

p_item_poly <- pheatmap(
  poly_mat, 
  color = coolwarm, 
  breaks = breaks_full,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Item Correlation (Polychoric)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

p_item_diff <- pheatmap(
  diff_item_mat, 
  color = coolwarm, 
  breaks = breaks_diff,
  cluster_rows = FALSE, 
  cluster_cols = FALSE,
  main = "Difference (Polychoric - Pearson)",
  show_rownames = FALSE, 
  show_colnames = FALSE, 
  silent = TRUE
)

# Display the heatmaps side by side
grid.arrange(
  p_item_pearson$gtable, 
  p_item_poly$gtable, 
  p_item_diff$gtable, 
  ncol = 3, 
  widths = c(1, 1, 1)
)

# Print a numerical summary of the differences in the console
cat("Numerical summary of the differences (Polychoric - Pearson) between items:\n")
summary(as.vector(diff_item_mat))
```

As we have said previously, the plot clearly indicates that Pearson
correlation tends to systematically underestimate the true correlation
between *questions* (not factor).

# Take the Big 5 personality test

You do not need to run all the code above to run this test. You only
need to run the cells before the factor analysis part and the first cell
from alternative B in order to take this test.

For more worked out tests you can go to the
[psytests.org](https://psytests.org/big5/ineoBen.html) website for 120
and 300 item questionnaires. It includes facets scores, which are
essentially subfactors of the Big 5 factors. For example the six facet
of openness are active imagination (fantasy), aesthetic sensitivity,
attentiveness to inner feelings, preference for variety
(adventurousness), intellectual curiosity, and challenging authority
(psychological liberalism).

Run this chunk interactively in your R console to answer the 50 IPIP
questions. Your Big 5 traits will be based on the previous EFA with
oblique rotation (for more realism and better interpretability), minres
method and polychoric correlation matrix.

Rate each statement from 1 (Disagree) to 5 (Agree).

```{r}
#| eval: false
questions <- c(
  EXT1="I am the life of the party.", EXT2="I don't talk a lot.", 
  EXT3="I feel comfortable around people.", EXT4="I keep in the background.", 
  EXT5="I start conversations.", EXT6="I have little to say.", 
  EXT7="I talk to a lot of different people at parties.", EXT8="I don't like to draw attention to myself.", 
  EXT9="I don't mind being the center of attention.", EXT10="I am quiet around strangers.",
  
  EST1="I get stressed out easily.", EST2="I am relaxed most of the time.", 
  EST3="I worry about things.", EST4="I seldom feel blue.", 
  EST5="I am easily disturbed.", EST6="I get upset easily.", 
  EST7="I change my mood a lot.", EST8="I have frequent mood swings.", 
  EST9="I get irritated easily.", EST10="I often feel blue.",
  
  AGR1="I feel little concern for others.", AGR2="I am interested in people.", 
  AGR3="I insult people.", AGR4="I sympathize with others' feelings.", 
  AGR5="I am not interested in other people's problems.", AGR6="I have a soft heart.", 
  AGR7="I am not really interested in others.", AGR8="I take time out for others.", 
  AGR9="I feel others' emotions.", AGR10="I make people feel at ease.",
  
  CSN1="I am always prepared.", CSN2="I leave my belongings around.", 
  CSN3="I pay attention to details.", CSN4="I make a mess of things.", 
  CSN5="I get chores done right away.", CSN6="I often forget to put things back in their proper place.", 
  CSN7="I like order.", CSN8="I shirk my duties.", 
  CSN9="I follow a schedule.", CSN10="I am exacting in my work.",
  
  OPN1="I have a rich vocabulary.", OPN2="I have difficulty understanding abstract ideas.", 
  OPN3="I have a vivid imagination.", OPN4="I am not interested in abstract ideas.", 
  OPN5="I have excellent ideas.", OPN6="I do not have a good imagination.", 
  OPN7="I am quick to understand things.", OPN8="I use difficult words.", 
  OPN9="I spend time reflecting on things.", OPN10="I am full of ideas."
)

my_answers <- numeric(length(questions))
names(my_answers) <- names(questions)

questions_order <- c(sapply(1:10, function(x) seq(x, 50, by = 10))) #shuffle questions
# Interactive prompt loop1
for (i in questions_order) {
  valid <- FALSE
  while (!valid) {
    ans <- readline(prompt = paste0(questions[i], " [1-5]: "))
    ans_num <- as.numeric(ans)
    
    if (!is.na(ans_num) && ans_num >= 1 && ans_num <= 5 && ans_num %% 1 == 0) {
      my_answers[i] <- ans_num
      valid <- TRUE
    } else {
      cat("Invalid input. Please enter an integer between 1 and 5.\n")
    }
  }
}

my_answers <- my_answers[colnames(data_mat)]

# Standardize and calculate factor scores
pop_means <- colMeans(data_mat)
pop_sds   <- apply(data_mat, 2, sd)
my_answers_std <- (my_answers - pop_means) / pop_sds

my_scores <- as.numeric(t(my_answers_std) %*% fa_obli$weights)
my_scores <- my_scores[c(1, 2, 3, 5, 4)]
names(my_scores) <- trait_names

# Calculate percentiles based on the original distribution
my_percentiles <- sapply(trait_names, function(t) {
  ecdf(data_fa_obli[[t]])(my_scores[t])
})

results <- data.frame(
  Factor_Score = round(my_scores, 3),
  Percentile = paste0(round(my_percentiles * 100, 1), " %")
)

print(results)
```
